---
title: "Gapminder exploration"
output: github_document 
keep_md: true 
---

## Bring the data in

```{r}
library(gapminder)
library(tidyverse)
```

## Smell test the data
- Is it a data.frame, a matrix, a vector, a list?
```{r}
typeof(gapminder)
```
It is a list.

- What’s its class?
```{r}
class(gapminder)
```
It is a data frame.

- How many variables/columns?
```{r}
ncol(gapminder)
```
It has 6 columns.

- How many rows/observations?
```{r}
nrow(gapminder)
```
It has 1704 rows.

- Can you get these facts about “extent” or “size” in more than one way? Can you imagine different functions being useful in  different contexts?

1. Alternative 1: 

Gives dimentions of dataframe in 1 command.
```{r}
dim(gapminder)
```

2. Alternative 2: 

Provides dimensions and type of each variable.
```{r}
str(gapminder)
```

- What data type is each variable?

Country and continent are factors (entered as integer dummy variables), year is an integer, life expectancy is numeric (double), population is an integer, and gdpPercap is numeric (double)
