---
title: "Writing functions and working with lists"
output: github_document 
keep_md: true 
---

## 2. Writing Functions

I'll stick with gapminder for this exercise.

```{r}
suppressPackageStartupMessages(library(tidyverse))
library(knitr)
library(gapminder)
```

Just to remind myself of what it looks like: 

```{r}
kable(head(gapminder))
```

**1. A linear model**

I will write a basic function to fit a linear model for the relationship between life expectancy and year. As suggested by Jenny in her [tutorial](http://stat545.com/block012_function-regress-lifeexp-on-year.html), it's definitely a good idea to force the intercept to start at the first year in the gapminder data (1952) not the first year ever (0)!  

```{r}
lm_fit <- function(mydata, offset=1952) {
  the_lm <- lm(lifeExp ~ I(year-offset), data=mydata)
    coef(the_lm)
}
```

Let's see if it works on gapminder:

```{r}
lm_fit(gapminder)
```

It seems like the function is working, but I'll do a quick sanity check.

```{r}
lm(lifeExp ~ I(year-1952), gapminder)
```

Looks good! Now what about using the function on a single country within the gapminder data?

```{r}
gapminder %>%
  filter(country=="France") %>%
      lm_fit()
```

That's good for now, but in the next exercise I'll automate this task in order to fit a linear model for each country without have to repeat the above command hundreds of times.

However, I would also like to be able to fit a couple different types of models, such as:

**2. A quadratic model**

I'll include a squared term for year to account for potential non-linearity in the change in life expectancy over time.

```{r}
quad_fit <- function(mydata, offset=1952) {
  the_quad <- lm(lifeExp ~ I(year-offset) + I((year-offset)^2), data=mydata)
    coef(the_quad)
}

quad_fit(gapminder)
```

**2. A robust regression model**

```{r}
suppressPackageStartupMessages(library(MASS))

robust_fit <- function(mydata, offset=1952, method="M") {
  the_robust <- MASS::rlm(lifeExp ~ I(year-offset), method=method, data=mydata)
    coef(the_robust)
}

robust_fit(gapminder)
```

This method produces very similar coefficients as the linear model, but robust regression using M-estimation will be less influenced by outliers in life expectancy, so there will probably be bigger differences for certain countries. 

Now I want to apply these functions to nested dataframes consisting of the countries in gapminder.

## 6. Working with nested dataframes